---
title: "Session 2 Workbook — A Real Analysis in R"
subtitle: "MSc Epidemiology · Intro to R & SAS"
author: "Your name here"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 7.5, fig.height = 4.2, dpi = 110
)
```

# How to use this workbook

Open this file in RStudio **inside the seminar RStudio Project**, so that the `data/`
folder sits beside it. Then:

* Run one chunk at a time with the green ▶ button, or <kbd>Ctrl</kbd>/<kbd>Cmd</kbd> + <kbd>Shift</kbd> + <kbd>Enter</kbd>
* Work through the `## YOUR TURN` chunks — they are deliberately incomplete
* **Knit** the whole thing at the end (the Knit button) to prove it runs top to bottom

If knitting fails but the chunks all ran, you have an ordering problem: something
later depends on an object you created out of sequence. That is exactly the bug
that reproducibility checks are meant to catch, and it is worth fixing.

```{r packages}
# install.packages(c("tidyverse", "broom", "janitor", "gtsummary"))  # once, if needed
library(tidyverse)
library(broom)
```

---

# 1 · Import and first look

```{r import}
births    <- read_csv("data/births.csv",    show_col_types = FALSE)
mothers   <- read_csv("data/mothers.csv",   show_col_types = FALSE)
screening <- read_csv("data/screening.csv", show_col_types = FALSE)

dim(births); dim(mothers); dim(screening)
```

```{r glimpse}
glimpse(births)
```

Three things are already wrong with this data. Find them:

```{r first-look}
count(births, sex)                  # 1. ?
colSums(is.na(mothers))             # 2. ?
c(rows = nrow(screening), infants = n_distinct(screening$infant_id))   # 3. ?
```

> **1.** `sex` is coded four ways (`F`, `M`, `f`, `m`).
> **2.** `mat_bmi` has missing values.
> **3.** Some infants have more than one screening specimen.

## YOUR TURN

How many births are preterm (`ga_weeks < 37`), and what percentage is that?

```{r yt-1}
# sum(...)
# 100 * mean(...)
```

---

# 2 · Clean and derive

```{r clean}
births_clean <- births |>
  mutate(
    sex      = toupper(sex),
    bw_kg    = birth_weight_g / 1000,
    preterm  = ga_weeks < 37,
    ga_group = factor(
      case_when(
        ga_weeks < 32 ~ "very preterm",
        ga_weeks < 37 ~ "preterm",
        ga_weeks < 42 ~ "term",
        .default      = "post-term"
      ),
      levels = c("term", "preterm", "very preterm", "post-term")
    )
  )

count(births_clean, ga_group)
count(births_clean, sex)
```

`janitor::clean_names()` does the equivalent tidying for *variable names* — useful when
data arrives from Excel with names like `Birth Weight (g)`:

```{r janitor, eval=FALSE}
births |> janitor::clean_names() |> names()
```

## YOUR TURN

Create `bw_cat` with levels `very low` (<1500 g), `low` (<2500 g) and `normal`,
then cross-tabulate it against `ga_group`.

```{r yt-2}
# births_clean |>
#   mutate(bw_cat = case_when(...)) |>
#   count(ga_group, bw_cat)
```

---

# 3 · The join that ruins analyses

This is the single most important section of the workshop.

```{r bad-join}
naive <- left_join(births_clean, screening, by = "infant_id")

c(births = nrow(births_clean), naive_join = nrow(naive))
```

The row count changed. Find out why:

```{r why}
naive |> count(infant_id) |> filter(n > 1) |> nrow()
```

Now do it properly — one screening record per infant:

```{r good-join}
first_spec <- screening |> filter(specimen_no == 1)

analysis <- births_clean |>
  left_join(first_spec, by = "infant_id") |>
  left_join(mothers,    by = "momid")

stopifnot(nrow(analysis) == nrow(births_clean))   # fail loudly if not
nrow(analysis)
```

`stopifnot()` is a **guard rail**. Put one after every join. If a future version of the
data breaks your assumption, the script stops instead of quietly producing a different answer.

Audit who didn't link:

```{r audit}
unscreened <- anti_join(births_clean, screening, by = "infant_id")
nrow(unscreened)

bind_rows(
  unscreened |> mutate(grp = "no screen"),
  semi_join(births_clean, screening, by = "infant_id") |> mutate(grp = "screened")
) |>
  group_by(grp) |>
  summarise(n = n(),
            mean_ga = mean(ga_weeks),
            mean_bw = mean(birth_weight_g),
            .groups = "drop")
```

## YOUR TURN

Are the unscreened infants different in a way that would bias a complete-case analysis?
Write one or two sentences below, as you would in a Limitations paragraph.

> *Your answer here.*

---

# 4 · Summarise

```{r summarise}
analysis |>
  group_by(site, sex) |>
  summarise(
    n           = n(),
    mean_bw     = mean(birth_weight_g),
    sd_bw       = sd(birth_weight_g),
    pct_preterm = 100 * mean(preterm),
    .groups     = "drop"
  ) |>
  arrange(desc(n))
```

## YOUR TURN

Median and IQR of `x17ohp` by `ga_group`, with the number of non-missing values in each group.
Remember `na.rm = TRUE`.

```{r yt-3}
# analysis |>
#   group_by(ga_group) |>
#   summarise(n = sum(!is.na(x17ohp)),
#             median = ...,
#             q1 = quantile(..., 0.25, na.rm = TRUE),
#             q3 = ...)
```

---

# 5 · Dates

```{r dates}
timing <- analysis |>
  mutate(
    birth_date  = as_date(birth_datetime),
    birth_month = month(birth_datetime, label = TRUE),
    hours_to_collection = as.numeric(
      difftime(collection_datetime, birth_datetime, units = "hours")
    )
  )

timing |>
  summarise(
    median_h = median(hours_to_collection, na.rm = TRUE),
    n_before_24h = sum(hours_to_collection < 24, na.rm = TRUE)
  )
```

---

# 6 · Missing data

```{r missing}
analysis |>
  summarise(across(everything(), ~ sum(is.na(.x)))) |>
  pivot_longer(everything(), names_to = "variable", values_to = "n_missing") |>
  filter(n_missing > 0) |>
  arrange(desc(n_missing))
```

`drop_na()` is one keystroke and a substantive assumption (MCAR). For anything going into
a manuscript, look at `mice::mice()` for multiple imputation.

---

# 7 · Table 1

Hand-rolled, so you can see what it is doing:

```{r table1-manual}
analysis |>
  mutate(preterm_lab = if_else(preterm, "Preterm (<37 wk)", "Term")) |>
  group_by(preterm_lab) |>
  summarise(
    n             = n(),
    mat_age       = sprintf("%.1f (%.1f)", mean(mat_age), sd(mat_age)),
    mat_bmi       = sprintf("%.1f (%.1f)", mean(mat_bmi, na.rm = TRUE), sd(mat_bmi, na.rm = TRUE)),
    smoking_pct   = sprintf("%.1f", 100 * mean(smoking == "yes")),
    hypertens_pct = sprintf("%.1f", 100 * mean(hypertension)),
    birthwt       = sprintf("%.0f (%.0f)", mean(birth_weight_g), sd(birth_weight_g)),
    .groups = "drop"
  )
```

And the two-line version you will actually use. `gtsummary` exports cleanly to Word:

```{r table1-gtsummary, eval=FALSE}
library(gtsummary)

analysis |>
  select(preterm, mat_age, mat_bmi, smoking, hypertension, diabetes_pre,
         birth_weight_g, sex, delivery_mode) |>
  tbl_summary(by = preterm, missing_text = "Missing") |>
  add_p() |>
  add_overall() |>
  bold_labels()
```

---

# 8 · Figures

```{r fig-hist}
ggplot(analysis, aes(ga_weeks)) +
  geom_histogram(binwidth = 0.5, fill = "#8b1d41", colour = "white") +
  geom_vline(xintercept = 37, linetype = "dashed", colour = "#12707a", linewidth = 0.9) +
  labs(x = "Gestational age (completed weeks)", y = "Infants",
       title = "Gestational age at birth") +
  theme_minimal(base_size = 13)
```

```{r fig-analyte}
analysis |>
  filter(!is.na(x17ohp)) |>
  ggplot(aes(ga_weeks, x17ohp)) +
  geom_point(alpha = 0.35) +
  geom_smooth(method = "loess", colour = "#8b1d41") +
  scale_y_log10() +
  labs(x = "Gestational age (weeks)", y = "17-OHP (nmol/L, log scale)",
       title = "17-OHP falls steeply with increasing gestational age") +
  theme_minimal(base_size = 13)
```

## YOUR TURN

Birth weight against gestational age, coloured by `sex`, with a linear smooth,
faceted by `site`. Give it publication-ready axis labels.

```{r yt-4}
# ggplot(analysis, aes(...)) + ...
```

---

# 9 · Models

```{r models}
m_bw <- lm(birth_weight_g ~ ga_weeks + sex + smoking + mat_age, data = analysis)
tidy(m_bw, conf.int = TRUE) |> mutate(across(where(is.numeric), \(x) round(x, 2)))
glance(m_bw) |> select(r.squared, adj.r.squared, sigma, nobs)

m_pt <- glm(preterm ~ mat_age + smoking + hypertension + diabetes_pre,
            data = analysis, family = binomial)
tidy(m_pt, exponentiate = TRUE, conf.int = TRUE) |>
  mutate(across(where(is.numeric), \(x) round(x, 3)))
```

A forest plot straight from `tidy()` output — this is why `broom` matters:

```{r forest}
tidy(m_pt, exponentiate = TRUE, conf.int = TRUE) |>
  filter(term != "(Intercept)") |>
  ggplot(aes(estimate, fct_rev(factor(term)))) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey50") +
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high)) +
  scale_x_log10() +
  labs(x = "Odds ratio (95% CI)", y = NULL, title = "Predictors of preterm birth") +
  theme_minimal(base_size = 13)
```

## YOUR TURN — think like an epidemiologist, not a programmer

Twins share a mother, so rows in `analysis` are **not independent**.

1. How many mothers contribute more than one infant?
2. Why does that matter for the standard errors in `m_pt`?
3. What would you do about it?

```{r yt-5}
# analysis |> count(momid) |> filter(n > 1) |> nrow()
```

> *Your answer here.* (Hint: cluster-robust variance, e.g. `sandwich` + `lmtest`,
> or a random intercept for `momid` with `lme4::glmer()`.)

---

# 10 · Spot the bugs

An AI assistant produced the code below when asked for "the analysis of preterm birth".
**It runs without error.** There are four things wrong with it. Fix them.

```{r bugs, eval=FALSE}
cohort <- births |>
  left_join(screening, by = "infant_id") |>
  left_join(mothers,   by = "momid")

cohort |> summarise(
  n          = n(),
  mean_bw    = mean(birth_weight_g),
  mean_17ohp = mean(x17ohp)
)

count(cohort, sex)

m <- glm(I(ga_weeks < 37) ~ mat_age + smoking + mat_bmi,
         data = cohort, family = gaussian)
tidy(m, exponentiate = TRUE)
```

```{r bugs-fixed}
# Your corrected version here.
```

<details>
<summary>Answers (open only after you've tried)</summary>

1. The first `left_join()` duplicates infants with a repeat specimen —
   `nrow()` goes from 1239 to 1372. Filter to `specimen_no == 1` first.
2. `mean(x17ohp)` returns `NA` because 17-OHP has missing values. Needs `na.rm = TRUE`
   — *and* you should report how many were missing.
3. `sex` is still coded `F`/`M`/`f`/`m`, so any sex-stratified result is wrong.
4. `family = gaussian` on a TRUE/FALSE outcome is a linear probability model;
   `exponentiate = TRUE` then produces a meaningless number. Use `family = binomial`.
5. (Bonus) Rows are clustered within `momid` because of twins.

</details>

---

# 11 · Reproducibility check

```{r session}
sessionInfo()
```

Knit this document. If it knits from a clean session, your analysis is reproducible.
If it doesn't, find out why now rather than the week before your defence.
